---
title: "Project 2 Solution"
author: "Ananya Bhaktaram"
class: "Statistical Computing"
date: "September 29th,2024"
output: html_document
---
```{r}
# Install necessary packages
install.packages("tidyverse")
install.packages ("tidytuesdayR")

# Load Respective packages
library(tidyverse)
library(tidytuesdayR)
```

# Part 1: Fun with functions

### Part 1A: Exponential transformation

The exponential of a number can be written as an infinite series expansion of the form $$
\exp(x) = 1 + x + \frac{x^2}{2!} + \frac{x^3}{3!} + \cdots
$$ Of course, we cannot compute an infinite series by the end of this term and so we must truncate it at a certain point in the series. The truncated sum of terms represents an approximation to the true exponential, but the approximation may be usable.

Write a function that computes the exponential of a number using the truncated series expansion. The function should take two arguments:

-   `x`: the number to be exponentiated

-   `k`: the number of terms to be used in the series expansion beyond the constant 1. The value of `k` is always $\geq 1$.

For example, if $k = 1$, then the `Exp` function should return the number $1 + x$. If $k = 2$, then you should return the number $1 + x + x^2/2!$.

Include at least one example of output using your function.

::: callout-note
-   You can assume that the input value `x` will always be a *single* number.

-   You can assume that the value `k` will always be an integer $\geq 1$.

-   Do not use the `exp()` function in R.

-   The `factorial()` function can be used to compute factorials.

```{r}
Exp <- function(x, k) {
    # Add your solution here
}
```

